CMSC818s: Project 3 \\ A Brief Survey of Modern Virtualization Research

### 2011-12-12 paper.tex
greg's stuff

\documentclass[10pt]{article}
\pagestyle{plain}

\usepackage[margin=2.5cm]{geometry}
\usepackage{multicol}
\usepackage{url}

@@

\maketitle

@@

\begin{multicols*}{2}

@@

The rest of the paper is structured as follows: in Section \ref{sec:background}
we provide background information on the history of virtualization and the
sorts of techniques which are utilized today.  Section \ref{sec:summaries}
presents summaries of three recent virtualization papers from the
9\textsuperscript{th} USENIX Symposium on Operating Systems Design
and Implementation, as well as a fourth from (TBD!).  Section
\ref{sec:relationships} analyses these papers in terms of how they
relate to each other and how they approach the larger problem space.
Finally, in Section \ref{sec:conclusion} we conclude by discussing our
thoughts and opinions on the problem space as a whole, and postulate
about how virtualization may evolve in the future.

@@

@@

Turtles is designed for the x86 architecture, which subscribes to a
single-level hardware support model.  This means that there are only two
levels of execution privilege.  Turtles runs its host hypervisor, which
we will call $L_0$, at the most privileged level of execution.  Now, suppose
we have a hypervisor running above $L_0$ at level $L_1$, and that hypervisor
is running a guest OS at level $L_2$.  Everything running at $L_1$ and $L_2$
has to run at the same privilege level, and only $L_0$ is privileged enough
to interact with the hardware.  Essentially, levels $L_1$ through $L_n$ only
exist logically, and in reality they are all multiplexed to be running directly
on top of $L_0$.  All traps generated at $L_2$ go immediately to $L_0$, which
may handle them itself, or forward them back to $L_1$ for handling if hardware
interaction is not required.

@@

The mClock algorithm works by assigning numerical tags to each queued request.
Essentially, tags for VM $i$ with proportional weight $w_i$ will be spaced by $1/w_i$.
Requests will be executed in the order of their assigned tags, which ensures that
VM $i$ receieves his proportional share.  This approach requires a global clock
for tag assignment, so that idle clients remain synced with active ones and do not
amass ``idle credit'', thus gaining preferential treatment.  The tag for VM $i$ and
request $j$ is set as follows:

@@

The paper proceeds to rigorously evaluate the performance of Xen Clocksource
using NTP against RADClock, both in dependent and independent modes of
operation, and in a variety of different system conditions.  The conclusion
seems to be that RADClock is generally unaffected by the errors and
drift that plague clocksource.  This is especially clear in a series of
experiments conducted involving VM migrations, where ``migration shock''
pushes the Independent Clocksource clock into a state of instability that
takes nearly 20 minutes to correct, while dependent RADClock converges
instantly.  Furthermore, the RADClock algorithm works equally well
regardless of the choice of underlying hardware counter, and is much
less affected by high contention for the network and system resources,
making it a better solution for the problem of timekeeping in the
context of virtualization.

\subsection{Paper 4 - TBD}
\label{sec:summaries/tbd}

\section{Paper Relationships}
\label{sec:relationships}

@@

@@

\bibitem{ref:xen}
	P. Barham, B. Dragovic, K. Fraser, S. Hand, T. Harris, A. Ho, R. Neugebauer,
	I. Pratt, A. Warfield.  Xen and the Art of Virtualization.  In {\it Proc. of
	the 19\textsuperscript{th} ACM Symp. on Operating Systems Principles}, 2003.

@@

\end{multicols*}

>>>>

\documentclass[letterpaper, twocolumn]{article}
\pagestyle{plain}

\usepackage[margin=2.5cm]{geometry}
\usepackage{multicol}
\usepackage{url}
\usepackage{graphicx}

@@

\twocolumn[
\begin{@twocolumnfalse}
\maketitle

@@

\end{@twocolumnfalse}
]

@@

The rest of the paper is structured as follows: in Section \ref{sec:background}
we provide background information on the history of virtualization and the
sorts of techniques which are utilized today.  Section \ref{sec:summaries}
presents summaries of a foundation virtualization paper from the
5\textsuperscript{th} USENIX Symposium on Operating Systems Design
and Implementation, as well as three recent papers from the virtualization
session of the same Symposium in 2010.  Section \ref{sec:relns} attempts
to analyze and classify these papers in terms of how they
relate to each other and how they approach the larger problem space.
Finally, in Section \ref{sec:conclusion} we conclude by discussing our
thoughts and opinions on the problem space as a whole, and postulate
about how virtualization may evolve in the future.

@@

\subsection{Scale and Performance in the Denali Isolation Kernel}
\label{sec:summaries/denali}

@@

Turtles is designed for the x86 architecture, which subscribes to a
single-level hardware support model.  This means that there are only two
levels of execution privilege.  Turtles runs its host hypervisor, which
we will call $L_0$, at the most privileged level of execution.  Now, suppose
we have a hypervisor running above $L_0$ at level $L_1$, and that hypervisor
is running a guest OS at level $L_2$.  Everything running at $L_1$ and $L_2$
has to run at the same privilege level, and only $L_0$ is privileged enough
to interact with the hardware.  Essentially, levels $L_1$ through $L_n$ only
exist logically, and in reality they are all multiplexed to be running directly
on top of $L_0$ (see Figure \ref{fig:turtles-multiplex}).  All traps generated at
$L_2$ go immediately to $L_0$, which
may handle them itself, or forward them back to $L_1$ for handling if hardware
interaction is not required.

\begin{figure*}[t]
	\begin{center}
		\includegraphics[width=4in]{images/turtles.png}
	\end{center}
	\caption{The Turtles hypervisor maintains logical nested virtualization, but
		in reality multiplexes all levels of nesting into $L_1$. (Figure courtesy
		of \cite{ref:turtles})}
	\label{fig:turtles-multiplex}
\end{figure*}

@@

\begin{figure}[t]
	\begin{center}
		\includegraphics[width=2.5in]{images/mclock.png}
	\end{center}
	\caption{The mClock algorithm works by tagging requests with a tag for
		reservations, weights, and reservations, and then executing requests
		in order of their tags.  (Figure courtest of \cite{ref:mclock})}
	\label{fig:mclock-alg}
\end{figure}

The mClock algorithm (see Figure \ref{fig:mclock-alg}) works by assigning numerical tags to each
queued request.
Essentially, tags for VM $i$ with proportional weight $w_i$ will be spaced by $1/w_i$.
Requests will be executed in the order of their assigned tags, which ensures that
VM $i$ receieves his proportional share.  This approach requires a global clock
for tag assignment, so that idle clients remain synced with active ones and do not
amass ``idle credit'', thus gaining preferential treatment.  The tag for VM $i$ and
request $j$ is set as follows:

@@

The paper proceeds to rigorously evaluate the performance of Xen Clocksource
using NTP against RADClock, both in dependent and independent modes of
operation, and in a variety of different system conditions.  The conclusion
seems to be that RADClock is generally unaffected by the errors and
drift that plague clocksource (see Figure \ref{fig:time-eval}).
This is especially clear in a series of
experiments conducted involving VM migrations, where ``migration shock''
pushes the Independent Clocksource clock into a state of instability that
takes nearly 20 minutes to correct, while dependent RADClock converges
instantly.  Furthermore, the RADClock algorithm works equally well
regardless of the choice of underlying hardware counter, and is much
less affected by high contention for the network and system resources,
making it a better solution for the problem of timekeeping in the
context of virtualization.

\begin{figure*}[t]
	\caption{The RADClock algorithm consistently outperforms ntpd clocksource;
		here, the drift in clocksource is simply due to feedback from increased
		latency in the network. (Figure courtesy of \cite{ref:time})}
	\label{fig:time-eval}
	\begin{center}
		\includegraphics[width=6in]{images/time.png}
	\end{center}
\end{figure*}

\section{Paper Relationships}
\label{sec:relns}

As we surveyed this series of papers, we noticed that the literature
in this area seems to neatly classify itself by \emph{motivation}.
Many papers on virtualization come from the realm of academia, and these
are mostly characterized by a focus on clean abstractions and solutions
with good theoretical properties, such as fairness and strong isolation.
Many more papers, however, originate from corporate research at places
like Microsoft and HP Labs.  These papers seem much more interested
in providing fast, scalable solutions using commodity hardware and
software.  In our papers, we found a series of tensions which seemed
to be expressing this fundamental question of motivation.  We will
now turn our attention to these tensions, keeping in mind that the
Turtles and mClock papers share corporate authorship, while the
Denali and Time papers have originated in the realm of academia.

\subsection{Performance and Fairness}
\label{sec:relns/perf-fair}

The most obvious tension which surfaced in the literature was a question
of priority between performance and fairness.  Generally speaking, there is a trend
in corporate papers to be heavily concerned with speed and efficiency, while
academically-minded papers seem more focused on strong properties of isolation
and guarantees of fairness between VMs.  This should not be particularly surprising,
given the authorship; more efficient virtualization seems to be an easily reachable
step toward greater profit.  It is worth noting, however, that in the more recent
literature corporate papers seem to be taking note of the importance of fairness;
we will discuss this further in Section \ref{sec:relns/scale}.

Our papers are clearly divided by this metric:  Turtles is all about efficiency
and performance, often at the cost of fairness.  The problem seems
especially hard here because of the nested context.  It seems perfectly logical that if
a guest OS running in $L_2$ should only receive a fraction $\alpha$ of the resources allocated
to its host in $L_1$, which in turn should only receive a percentage $\beta$ of the machine's
total resources, then these two fractions should compound and our guest should end up with
$\alpha\beta$ of the total capacity available.  However, in reality Turtles multiplexes
all guests at all levels onto $L_0$, so in many ways the guest OS and its hypervisor will
be treated as siblings rather than a parent and child.  Multi-level device assignment
for Port I/O and Memory-mapped I/O is particularly agnostic to these concerns, because
it works exactly the same way as single-level device assignment, and so doesn't care whether
a guest runs at $L_1$ or $L_{100}$.  Furthermore, no experimentation at all is done to analyze
the effects of this model on fairness, and the authors even make the statement that this
is a secondary concern, and so they are content with a sort of best-effort.

On the other hand, Denali and mClock are all about prioritizing fairness.  The foundational
goal of the Denali isolation kernel is to ensure strong isolation between clients, so that
the behavior of any particular VM cannot in any way affect another.  And the whole point of
the mClock scheduling algorithm is to ensure that limits, reservations, and weights
are met fairly between VMs contending for a resource.  It should be noted that neither
of these papers particularly eschew performance as a worthy goal; rather, the view
here seems more to be that fast performance for a particular guest doesn't really
mean much if a malicious (or even just selfish neighbor) can somehow choke that VM
out and make it impossible to make progress.

The Time paper is somewhat agnostic to concerns of performance and fairness anyway,
since it deals more with the abstractions presented to client applications.  We
will turn to this next.

\subsection{Paravirtualization and Full Virtualization}
\label{sec:relns/para-full}

The second tension seems to arise between the paradigms of \emph{paravirtualization}
and \emph{full virtualization}, which were introduced in Section \ref{sec:background}.
Corporately-funded literature seems to have an eye more toward using as much
commodity software as possible here, for the sake of lower adoption costs, while
academic papers seem more willing to consider modifying client software in
appropriate situations.

The split here is pretty clear as well: Denali, Xen, and Time all consider
paravirtualization to be beneficial and even necessary in the right situations.
In Denali, the isolation kernel presents a modified and simplified hardware abstraction
to its clients, which cleverly allows for some big wins in efficiency and performance
as a side effect.  Xen claims that paravirtualization is necessary to ensure fairness
and prevent starvation in process and I/O scheduling; one could argue, however, that
this is no longer a concern due to the advent of newer virtualization-aware hardware
(see Section \ref{sec:relns/hard-emul}).  The Time paper goes a step further and claims that
even with the new hardware, making the guest OS aware of the virtualized context it
runs in is necessary to ensure the correctness of certain time-dependent applications!

In contrast, the Turtles project argues for full virtualization and unmodified guest
OSes and hypervisors.  In the paper, the argument for this goes back to the
motivating question of efficiency, though it was unclear to us why paravirtualization
would have any negative impact here.  But it seems also that for any sort of large-scale
commercial deployment of server virtualization technology, the ability to use commodity
guests would be crucial in order to make adoption easy, and we consider this to also
be a likely factor influencing the design choice.

The mClock paper doesn't really fit into these two classifications, since it mostly
deals with the interactions between the hypervisor and its devices, rather than the hypervisor
and its guests.  However, the paper does mention that one of the reasons this disk
scheduling problem is so difficult is because guest OSes have their own algorithms
for efficient disk scheduling, which generally don't play nice with each other.  If it
were possible to modify the guest OS to disable these algorithms, so that the hypervisor
could handle all disk scheduling decision, the difficulty of this problem could likely
be reduced.

\subsection{Hardware Support and Software Emulation}
\label{sec:relns/hard-emul}

The ``tension'' between hardware support for virtualization and the use of software
emulation is perhaps better just described as a paradigm shift.  It seems
that over the last decade virtualization-friendly hardware has become commonplace
and inexpensive.  When the Xen and Denali papers were written, there was no real
alternative, and so everything had to be done through slow, inefficient software emulation.
But when hardware manufacturers began to produce virtualization-enabled chipsets
like the x86, this problem essentially disappeared overnight, and it has largely
remained out-of-scope for most of the intervening decade.

However, the question of software emulation may need to be re-examined as virtualization
research moves foward.  While hardware-enabled approaches are clearly much faster
than any attempt to do all the work in software, it seems that it just may not be possible to do
everything at the hardware level.  The Turtles project is a perfect example of this - there
is no hardware support for nested levels of virtualization, and so a lot of cleverness
is required to be able to compensate in the software without an unacceptable performance
penalty.  We have essentially been sent back to where single-level virtualization
solutions were a decade ago.  Some of the future work mentioned in this paper involved
examining ways to provide architectural support for nested virtualization, but although
we are somewhat skeptical that this will prove fruitful.  Attempting to provide
support for an unbounded number of levels of nesting with finite hardware does
not seem like it would be particularly successful.  It will be interesting to
see what direction this sort of work takes in the future, and whether hardware
designers will rise to the challenge of nested virtualization.

\subsection{Scalability and the Cloud}
\label{sec:relns/scale}

The final tension we wish to consider is the question of scalability.  We have observed
that there is a commercial interest in developing super-scalable server virtualization,
largely because it has found such a unique niche with the market growth of cloud computing
and infrastructure-as-a-service (IAAS).  While academic research is not uninterested in the question
of scale or the cloud use case, we do not feel that they are as much of a driving factor in this realm.

We can observe this trend in the papers we have been discussing.  The Turtles paper presents
the motivation for nested virtualization using the argument of its suitability for
infrastructure-as-a-service, as the two are essentially made for each other.  The mClock paper,
although it does not particularly market itself
for scalability, is motivated by the cloud-enabled scenario of many virtualized clients
needing to schedule the use of a shared network disk.  Furthermore, the guarantees mClock
makes about fairness, reservations, and proportional share are especially important
in this scenario, as they allow IAAS providers to keep service-level agreements with
their clients.  We find this particularly interesting, as it seems to indicate
that in the cloud, fairness and performance may find themselves on equal footing in
terms of importance (see Section \ref{sec:relns/perf-fair}).

Interestingly, the Denali paper also has a particular eye to scalability, and a large
portion of the evaluations section is devoted to demonstrating that the isolation kernel
performs well even when supporting over 10,000 VMs.  However, even in Denali, scalability
is just one goal amongst many, and the paper does not really consider applications
to cloud computing, being several years too early.

The Time paper does not really consider the question of scalability at all, being
somwhat orthogonal to the problem of accurate timekeeping on a single VM.  The paper
does point out that there are scalability benefits to the dependent RADClock algorithm,
as each VM no longer needs to send out synchronization packets on the network, but can
depend on the hypervisor to do this for all VMs that it hosts.  However, this is 
in many ways just an added bonus, and is not really a motivational argument for the paper.

@@

Having examined the recent literature along these four axes of tension, it is our belief
that a clear trend has emerged in virtualization.  We believe that as cloud computing and
IAAS continue to take off, research in this area will continue to trend toward what
is both popular and financially successful.  We expect that the majority of virtualization
research will continue in the direction of full virtualization over paravirtualization,
and that the cloud-computing/IAAS scenario will become the driving use case.
However, we think that as a result of this, fairness and performance will no longer be
competing ends, but will in fact become equal players.  This will be necessary in order
for cloud providers to meet their SLAs to their consumers, and we believe fair-scheduling
and resource-isolation research will move from its place in the shadows to the forefront
of both corporate and academic research.

As to the question of hardware support versus software emulation, we believe that as
new directions like nested virtualization are explored, some sort of hybrid approach
will continue to be needed.  However, we do not expect the status quo of fast hardware-enabled
virtualization to be shaken up in the general case; it seems that performance is too critical
in the driving cloud-based use cases, and we would be surprised to see nested virtualization
(or any other software-hybrid approach) gaining momentum here anytime soon.

Cynical though it may be, we feel that a lot of the interesting problems in virtualization
have already been solved, at least from an academic perspective.  So much of the recent
literature seems to be in the direction of just making things more efficient and more
reliable for the sake of the commercialized cloud.  We have stopped asking the question,
``can we do crazy thing $x$?'', and now are only asking ``how much faster can we make $x$ work?''
and ``how can we make $x$ work with fewer resources?''
This is certainly an important part of the research process for any area, but it is ostensibly
less exciting than the ground-breaking work that was going on earlier in the decade.

Of course, we cannot say with any certainty that there is not some new, groundbreaking discovery
just waiting to be made around the corner.  Such a discovery could revitalize the field
and take virtualization off into all sorts of new and exciting directions.  But we do feel that
the way the problem has been framed makes it unlikely that such a new discovery could shift
the focus away from cloud-based server virtualization.  With so much focus on making the problem
fit into the market niche, such a new discovery would have to be truly monumental indeed -
monumental enough that no one could possibly predict it.

@@

\bibitem{ref:xen}
	P. Barham, B. Dragovic, K. Fraser, S. Hand, T. Harris, A. Ho, R. Neugebauer,
	I. Pratt, and A. Warfield.  Xen and the Art of Virtualization.  In {\it Proc. of
	the 19\textsuperscript{th} ACM Symp. on Operating Systems Principles}, 2003.

\bibitem{ref:denali}
	A. Whitaker, M. Shaw, and S. Gribble.  Scale and Performance in the Denali Isolation
	Kernel.  In {\it Proc. of the 5\textsuperscript{th} Symp. on Operating Systems
	Desgin \& Implementation}, 2002.

@@
